
Using functional logic programing we can solve difficult problems in a very small amount of code.
As an example, we can solve the n-queens problem, in just 5 readable lines of code.
We can write simple declarative programs for many complex tasks.
Unfortunately, many computer scientists consider current implementations of functional logic languages
too slow to use in practice.
This has led to the idea that functional logic languages are inherently slow.
I believe a more likely culprit is the fact that very little work on program optimization 
has been incorporated into modern functional logic compilers.
In this dissertation I intend to develop an optimization framework for the functional logic language Curry.
This will include an intermediate language suitable for optimization;
implementations of the deforestation, unboxing, and shortcutting optimizations;
justification of the correctness of the implementations;
a compiler to a low level language;
and analysis of the running time and memory usage on several Curry programs.
I will show that it is possible to apply the optimizations of unboxing, deforestation, and shortcutting to Curry programs.
Furthermore, these optimizations will improve the performance of Curry programs.

Functional logic programming is a very powerful technique for expressing complicated ideas in a simple form.
Curry implements these ideas with a clean, easy to read syntax.
It is syntactically very similar to Haskell, a well known functional programming language.
It is also lazy, so evaluation of Curry programs is similar Haskell as well.
Curry extends Haskell with two new concepts.
First, there are non-deterministic functions, such as ``\texttt{?}''.  
Semantically \texttt{a} \texttt{?} \texttt{b} will evaluate \texttt{a} and \texttt{b} and will return both answers to the user.
Second, there are free, or logic, variables.
A free variable is a variable that is not in the scope of the current function.
The value of a free variable is not defined, but it may be constrained.

Consider the following Curry code for solving n-queens:
\begin{verbatim}
queens = isEmpty (set1 unsafe p) = p
 where p = permute [1..n]

unsafe qs
 | qs =:= (xs++[a]++ys++[b]++zs) = abs (a-b) =:= length ys
  where xs, a, ys, b, zs free
\end{verbatim}
In the \texttt{unsafe} function I've broken the list \texttt{qs} into 5 pieces.
two of the pieces, \texttt{a} and \texttt{b}, are lists with a single element.
the sublists, \texttt{xs}, \texttt{ys}, and \texttt{zs} are free to be as long as they want.
However, I've constrained the total list \texttt{(xs++[a]++ys++[b]++zs)}
to be the same as our argument \texttt{qs}.
the effect is that \texttt{a} and \texttt{b} are arbitrary elements in the list \texttt{q},
and \texttt{ys} is the list of elements between \texttt{a} and \texttt{b}.

Free variables are given concrete values in Curry programs through narrowing.
The semantics of narrowing and non-determinism in Curry are given by Antoy et al. \cite{Antoy97ALP}

There are currently two mature Curry compilers, Pakcs and Kics2.
Pakcs compiles Curry to Prolog in an effort to leverage Prolog's non-determinism and free variables.
Kics2 compiles Curry to Haskell in an effort to leverage Haskell's higher order functions and optimizing compiler.
Both compilers have their advantages.  Pakcs tends to perform better on non-deterministic expressions with free variables,
where Kics2 tends to perform much better on deterministic expressions.
Unfortunately neither of these compilers perform well in both circumstances.

Sprite, an experimental compiler, aims to fix these inefficiencies.
The strategy is to compile to a virtual assembly language, known as LLVM.
So far Sprite has shown promising improvements over both Pakcs and Kics2 in performance,
but it is not a mature compiler.

One major disadvantage of all three compilers is that they all attempt to pass off optimization to another compiler.
Pakcs attempts to have Prolog optimize the non-deterministic code; Kics2 attempts to use Haskell to optimize
deterministic code; and Sprite attempts to use LLVM to optimize the low level code.
Unfortunately none of these approaches works very well.
While some implementations of Prolog can optimize non-deterministic expressions, they have no concept of higher order functions,
so there are many optimizations that cannot be applied.
Kics2 is in a similar situation.  
In order to incorporate non-deterministic computations in Haskell, 
a significant amount of code must be threaded through each computation.
This means that any non-deterministic expression cannot be optimized in Kics2.
Finally, since LLVM doesn't know about either higher order functions or non-determinism, it loses many easy opprotunities for optimization.

Curry programs have one last hope for efficient execution.
Recently, many scientists \cite{peval_bjorn, offline_peval_Ramos} 
have developed a strong theory of partial evaluation for functional logic programs.
While these results are interesting, partial evaluation is not currently automatic in Curry.
Guidance is required from the programmer to run the optimization.
Furthermore, the optimization fails to optimize several common programs.

So far all of these approaches have failed to include into their implementations the large body of work on program optimizations.
This leads to the inescapable conclusion that Curry needs an optimizer.
The purpose of this proposal is, first, to make a case that Pakcs and Kics2 both fail to do an adequate job
of optimizing Curry programs, then to propose a new solution of building optimization into the Curry compiler pipeline.

This may seem trivial at first.  After all, optimization has been studied heavily for decades, and there are many well understood
optimizations for imperative, functional, and logic programming. 
\cite{optminzation_allen, dataflow_allen, LowryMedlock69, dataflow_kildall,
AhoUllman77, continuations_appel, compilers_appel, orbit, ssa_alpern, ssa_Wegman, ssa_wolfe,
steele78, stg-peytonJones, anormal_Flanagan, lambda_rename_steel, lambda_goto,
deforestation_wadler, shortcut_deforestation, haskell_inliner}
Why don't I just use these optimizations in Curry?
Ideally I can, but it's not immediately clear that these optimizations are valid in Curry.
This dissertation will contain two components: implementing some optimizations in Curry,
and proving the correctness of these optimizations.

I'm limiting myself to implementing deforestation, unboxing, and shortcutting in this dissertation.
While there are many other optimizations that I could implement, I believe that these will be the most beneficial.

The rest of this paper is organized as follows.
Section 2 presents the Curry language. Section 3 presents the need for an optimizing compiler in Curry.
Section 4, 5, and 6 present the history of compiler development for functional, logic, and functional logic languages respectively.
Section 7 presents the approaches to compiling Curry specifically.
Section 8 presents the optimizations I plan to implement.
Section 9 presents my final deliverable at the end of the dissertation.
Finally section 10 concludes.
